---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _datadelay_: Calculate robust epidemic case fatality rates

<!-- badges: start -->
[![R-CMD-check](https://github.com/epiverse-trace/datadelay/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/datadelay/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of datadelay is to provide robust estimates of the case fatality rate of an epidemic.

## Installation

You can install the development version of datadelay from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("epiverse-trace/datadelay")
```

## Quick start

This example of basic usage shows how to use _datadelay_ to estimate case fatality ratios from the 1976 Ebola outbreak.

```{r}
# Load package
library(datadelay)

# Load epiparameter for delay distributions
library(epiparameter)

# Load the Ebola 1976 data provided with the package
data("ebola1976")

# Get the onset to death distribution for ebola virus disease
onset_to_death_ebola <- epidist(
  pathogen = "ebola",
  delay_dist = "onset_to_death"
)
# Access the probability mass function
delay_pmf <- onset_to_death_ebola$pmf

# Calculate the static naive and corrected CFRs
ncfr <- static_cfr(ebola1976, correct_for_delays = FALSE)
ccfr <- static_cfr(ebola1976, correct_for_delays = TRUE, onset_to_death_ebola)

# Print nicely formatted case fatality rate estimates
format_cfr_neatly(ncfr)
format_cfr_neatly(ccfr)

# Calculate rolling naive and corrected CFRs
df_ncfr <- rolling_cfr(ebola1976, correct_for_delays = FALSE)
df_ccfr <- rolling_cfr(
  ebola1976,
  correct_for_delays = TRUE,
  onset_to_death_ebola
)

# Plotting case and death data along with CFRs
plot_data_and_cfr(df_ncfr, df_ccfr)
```
