---
output: github_document
bibliography: vignettes/resources/library.json
link-citations: true
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# _{{ packagename }}_: Estimating disease severity and under-reporting

<!-- badges: start -->
[![License: MIT](https://img.shields.io/badge/License-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![R-CMD-check](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/{{ gh_repo }}/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/{{ gh_repo }}/branch/main/graph/badge.svg)](https://app.codecov.io/gh/{{ gh_repo }}?branch=main)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![CRAN status](https://www.r-pkg.org/badges/version/{{ packagename }})](https://CRAN.R-project.org/package={{ packagename }})
<!-- badges: end -->

_{{ packagename }}_ is an R package to estimate disease severity and under-reporting in real-time, accounting for delays in epidemic time-series.

_{{ packagename }}_ provides simple, fast methods to calculate the overall or static case fatality ratio (CFR) of an outbreak up to a given time point, as well as how the CFR changes over the course of the outbreak.
_{{ packagename }}_ can help estimate disease under-reporting in real-time, accounting for delays reporting the outcomes of cases.

_{{ packagename }}_ implements methods outlined in @nishiura2009, and CFR estimates based on more methods are likely to be added.

_{{ packagename }}_ relies on the [_epiparameter_ package](https://epiverse-trace.github.io/epiparameter/), and both are developed at the [Centre for the Mathematical Modelling of Infectious Diseases](https://www.lshtm.ac.uk/research/centres/centre-mathematical-modelling-infectious-diseases) at the London School of Hygiene and Tropical Medicine as part of the [Epiverse-TRACE initiative](https://data.org/initiatives/epiverse/).

## Installation

The current development version of _{{ packagename }}_ can be installed from [GitHub](https://github.com/) using the `pak` package.

```r
if(!require("pak")) install.packages("pak")
pak::pak("{{ gh_repo }}")

# Also install R package {epiparameter} for epidemiological parameter values
pak::pak("epiverse-trace/epiparameter")
```

## Quick start

### Overall severity of the 1976 Ebola outbreak

This example of basic usage shows how to use _{{ packagename }}_ to estimate the overall case fatality ratios from the 1976 Ebola outbreak.

```{r example-ebola-calc}
# Load package
library(datadelay)

# Load the Ebola 1976 data provided with the package
data("ebola1976")

# read delay distribution for ebolavirus onset to death from {epiparameter}
# accesses parameters reported in https://doi.org/10.1016/S0140-6736(18)31387-4
onset_to_death_ebola <- epiparameter::epidist_db(
  disease = "Ebola Virus Disease",
  epi_dist = "onset_to_death",
  author = "Barry_etal"
)

# Calculate the static CFR without correcting for delays
estimate_static(data = ebola1976)

# Calculate the static CFR while correcting for delays
estimate_static(
  data = ebola1976,
  correct_for_delays = TRUE,
  epi_dist = onset_to_death_ebola
)
```

### Change in disease severity over the 1976 Ebola outbreak

Calculate the time-varying CFR estimates up to a given point in time.

```{r example-ebola-plot}
# Calculate the CFR without correcting for delays on each day of the outbreak
head(
  estimate_time_varying(
    data = ebola1976, correct_for_delays = FALSE,
    burn_in_value = 7
  )
)

# Calculate the daily CFR while correcting for delays
head(
  estimate_time_varying(
    ebola1976,
    correct_for_delays = TRUE,
    epi_dist = onset_to_death_ebola,
    burn_in_value = 7
  )
)

# plots to be added later
```

## Package vignettes

More details on how to use _{{ packagename }}_ can be found in the [online documentation as package vignettes](https://epiverse-trace.github.io/{{ packagename }}/), under "Articles".

## Help

To report a bug please open an [issue](https://github.com/{{ gh_repo }}/issues/new/choose).

## Contribute

Contributions to _{{ packagename }}_ are welcomed. Please follow the [package contributing guide](https://github.com/{{ gh_repo }}/blob/main/.github/CONTRIBUTING.md).

## Code of conduct

Please note that the _{{ packagename }}_ project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

## References
