---
title: "Estimating the proportion of cases that are reported during an outbreak"
output: rmarkdown::html_vignette
bibliography: resources/library.bib
csl: resources/bmj.csl
vignette: >
  %\VignetteIndexEntry{Estimating the proportion of cases that are reported during an outbreak}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- removes the ugly border around the figures -->
<style>
    img {
        border: 0;
    }
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align  = "left",
  fig.width  = 600/80,
  fig.height = 400/80,
  dpi = 80,
  fig.retina = 2,
  dev.args = list(type = "cairo-png")
)
```

# Overview

This vignette outlines the function within `datadelay` used to estimate the
proportion of cases, infections or hospitalisations ascertained.

We load the three packages required in this vignette using the following 
commands:

Now we load the five key packages we need to run the commands in this vignette:

```{r, message = FALSE, warning=FALSE, eval = TRUE}
library(datadelay)
library(epiparameter)
library(covidregionaldata)
library(knitr)
library(kableExtra)
```

and the four optional packages, to run the multiple examples in the last 
section of this vignette:

```{r, message = FALSE, warning=FALSE, eval = TRUE}
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
```


# Motivation

There are several reasons why cases, infections or hospitalisations and even 
deaths might be under-ascertained during an outbreak of an infectious disease. 
Some examples include:

* Testing effort or ability is not high enough to catch all infections.

* Testing triggered by individuals rather than sampling at random in the 
population. This can lead to biases due to some infections either being 
asymptomatic or having a pre-symptomatic phase. For example, if testing is 
initiated by an individual who appears not to have any symptoms, it is likely 
the ascertainment of their infection will be delayed, or possibly missed 
entirely.

* Changes in the criteria, either used to diagnose the disease in question or
to determine whether a death can be atttributed to the disease in question.

For example, during the recent COVID-19 pandemic, the U.K. were not able to roll
out widespread testing until around May 2020. This means that during March and 
April, infections were significantly under-ascertained. We focus on this period
within this vignette, as it illustrates the use for this analysis pipeline well.

# Methods

## Data required

The data required to estimate how the severity of a disease changes over time
using the `datadelay` package includes:

* A time-series of cases, hospitalisations or some other proxy for infections 
over time;
* A time-series of deaths;
* A delay distribution, describing the probability an individual will die $t$ 
days after they were initially exposed. Such distributions come from the
literature, where studies have typically fit distributions to data describing 
the process. 

In practice, the time-series of cases and deaths will already be together, given
that they usually originate from sources that will have typically collated them
into a single data file.

The package requires a `data.frame` of input data --- typically case and death 
time series data --- and a delay distribution. The delay distribution we use 
here comes from the `epiparameter` package.

## Estimating the proportion of cases or infections that have been ascertained

The function `estimate_reporting()` from the `datadelay` package estimates the 
proportion of cases, infections, hospitalisations --- or whichever proxy for 
infections is provide --- which have been ascertained. The method used within 
this function extends the methods outlined in the previous vignettes about 
estimating the severity during an ongoing outbreak and measuring how the 
severity changes over time. Both methods are based on the Nishiura et 
al.^[@nishiura2009early]^ methods to estimate severity. We extend this by 
combining the resulting severity estimates with an assumed baseline severity 
estimate. The proportion of cases, infections or other quantity provided that 
have been ascertained is given by the ratio of the assumed to be true baseline 
severity estimate to the delay-adjusted severity estimate. The delay-adjusted 
severity estimates can be calculated using either the `estimate_static()` or the
`estimate_time_varying()` functions. 

# Example with data from the ongoing COVID-19 pandemic in the U.K.

We outline how the time-varying severity estimation works in *datadelay* using 
a number of examples. The data for all of the examples is from the ongoing 
COVID-19 epidemic. Firstly, we analyse the U.K. data, then we pick three other 
countries with large outbreaks to analyse. 

## Downloading the raw data

First of all, we subset the data so that we focus on just the first few months 
of the COVID-19 outbreak in the U.K. We do so, as test availability changed 
dramatically as a result of the vaccine campaign. We download the data — 
using the `covidregionaldata` package — with the following 
command:

```{r, message = FALSE, warning = FALSE, eval = TRUE}
df_covid_uk <- get_national_data(
  countries = "united kingdom", source = "who", verbose = FALSE) |>
  dplyr::rename(cases = cases_new, deaths = deaths_new)
```

We then subset the data to focus on just the first few months of the outbreak 
with the following command:

```{r, message = FALSE, warning = FALSE, eval = TRUE}
df_covid_uk_subset <- subset(df_covid_uk, date <= "2020-05-31")
```

## Plotting the raw data

First, we plot case incidence data with following command:

```{r, message = FALSE, warning = FALSE, eval = TRUE, fig.cap = "\\label{fig:raw_data_covid} **Figure 1:** Incidence of cases over time for the ongoing COVID-19 outbreak in the U.K."}
plot_case_data(df_covid_uk_subset)
```

Then, we plot the death incidence data with following command:

```{r, message = FALSE, warning = FALSE, eval = TRUE, fig.cap = "\\label{fig:raw_data_covid} **Figure 2:** Incidence of  deaths over time for the ongoing COVID-19 outbreak in the U.K."}
plot_death_data(df_covid_uk_subset)
```

## The delay distribution

We again retrieve the appropriate distribution — reported 
[here](https://doi.org/10.3390/jcm9020538)^[@linton2020incubation]^ — using the
`epidist_db()` function from the `epiparameter` package, using the following
command:

```{r message=FALSE, warning=FALSE, eval=TRUE}
onset_to_death_covid <- epidist_db(
  disease = "COVID-19",
  epi_dist = "onset_to_death",
  author = "Linton_etal"
)
```

To visualise this distribution, we evaluate it between 0 and 30 days, and plot
the results over time. We do so using the following command:

```{r message=FALSE, warning=FALSE, eval=TRUE, fig.cap = "\\label{fig:delay_dist_covid} **Figure 3: Example plot of the appropriate delay distribution for the COVID-19 epidemic in the U.K.** We plot the onset-to-death distribution we use throughout this example for COVID-19, reported in https://doi.org/10.3390/jcm9020538."}
plot_epiparameter_distribution(onset_to_death_covid, by = 0.1)
```

## Estimating the proportion of cases that have been ascertained

We use the `estimate_reporting()` function within the `datadelay` package to 
calculate the time-varying CFR for the COVID-19 epidemic in the U.K:

```{r message = FALSE, warning = FALSE, eval = TRUE}
df_reporting_static <- estimate_reporting(df_covid_uk_subset,
                                          epi_dist = onset_to_death_covid,
                                          type = "static",
                                          severity_baseline = 0.014,
                                          correct_for_delays = TRUE) |> 
  format_output(estimate_type = "reporting", type = "Under-reporting")

```

The function includes a `type` argument, which determines whether
`estimate_static()` or `estimate_time_varying()` is used to estimate the 
delay-adjusted severity of the disease used as the numerator in the
under-ascertainment calculation. The `severity_baseline` argument in the
`estimate_reporting()` determines the denominator in the resulting 
under-ascertainment calculation. The other arguments are the same as those 
found in the `estimate_time_varying()`. 

```{r message = FALSE, warning = FALSE, eval = TRUE}
df_reporting_varying <- estimate_reporting(df_covid_uk,
                                           epi_dist = onset_to_death_covid,
                                           type = "varying",
                                           severity_baseline = 0.014,
                                           smooth_inputs = TRUE,
                                           burn_in = TRUE,
                                           correct_for_delays = TRUE,
                                           custom_date = "2020-06-30") |> 
  format_output(estimate_type = "reporting", type = "Under-reporting")
```

Once we have calculated both severity quantities over time, we plot the results.
First we plot the naive (uncorrected for delays) estiamte:

```{r message=FALSE, warning=FALSE, eval=TRUE}
knitr::kable(df_reporting_static,
  align = "c",
  col.names = c("Location", "Under-reporting estimate", "Type")) |>
  kableExtra::kable_styling(full_width = FALSE)
```

```{r message=FALSE, warning=FALSE, eval=TRUE}
knitr::kable(df_reporting_varying,
  align = "c",
  col.names = c("Location", "Under-reporting estimate", "Type")) |>
  kableExtra::kable_styling(full_width = FALSE)
```


# Example with all countries with large early COVID-19 outbreaks

```{r, message = FALSE, warning = FALSE, eval = TRUE}
df_covid <- get_national_data(
  source = "who", verbose = FALSE) |>
  rename(cases = cases_new, deaths = deaths_new)

df_covid_subset <- subset(df_covid, date <= "2020-05-31")
```

Finally, we put the ascertainment estimates for all countries with large 
outbreaks in a single table and summarise the results in a single figure. We 
define a large outbreak as one which has caused at least 100,000 deaths. We 
focus on these countries as the resulting table and figure become unwieldy if
they are made any larger. We focus on the period between the start of each 
outbreak to the 1st June 2020. We do so as it was this period where 
under-ascertainment of cases and infections was likely the highest, as tests
were still being developed and had not been made widely available in many 
countries.

To calculate all ascertainment estimates for all countries with
large COVID-19 outbreaks up to the 1st June 2020, we use the following command —
this codeblock depends on the package `dplyr`, for ease with manipulating
large data.frames which are structured by group:

```{r message = FALSE, warning = FALSE, eval = TRUE}
df_reporting <- df_covid |> 
  group_by(iso_code) |>
  mutate(total_deaths = max(deaths_total)) |> 
  filter(total_deaths > 100000 & !is.na(country)) |>
  filter(date < "2020-06-01") |>
  group_map(~estimate_reporting(
    df_in = .,
    epi_dist = onset_to_death_covid,
    type = "varying",
    severity_baseline = 0.014,
    burn_in = TRUE,
    smooth_inputs = TRUE,
    correct_for_delays = TRUE)) |> 
  bind_rows()

df_reporting_neat <- format_output(
  df_in = df_reporting,
  estimate_type = "reporting", 
  type = "Ascertainment")
```

Then, to plot all of these results on a single figure, we use the following 
command — this codeblock depends on the commonly used package `ggplot2`, for
ease with plotting large data.frames:

```{r message = FALSE, warning = FALSE, eval = TRUE, fig.cap = "\\label{fig:time_varying_cfr_corrected} **Figure 5: Example plot of the corrected time-varying CFR.** We calculate the time-varying CFR for the ongoing COVID-19 epidemic in United States, corrected for delays."}
df_reporting |> 
  ggplot() + 
  geom_pointrange(
    aes(x = fct_reorder(location, reporting_me), 
        y = reporting_me,
        ymin = reporting_lo, 
        ymax = reporting_hi,
        colour = location)) + 
  coord_flip() + 
  labs(x = NULL, y = "Ascertainment Rate") + 
  theme_minimal() + 
  theme(legend.position = "none") + 
  scale_y_continuous(labels = percent, limits = c(0, 1))
```

Finally, we put the same results in a large table with the following command:

```{r message = FALSE, warning = FALSE, eval = TRUE}
kable(df_reporting_neat,
  align = "c",
  col.names = c("Location", "Estimate", "Type")) |>
  kable_styling(full_width = FALSE)
```

