---
title: "Estimating fatality risk from individual level data"
output:
  bookdown::html_vignette2:
    fig_caption: yes
    code_folding: show
bibliography: resources/library.json
link-citations: true
vignette: >
  %\VignetteIndexEntry{Estimating fatality risk from individual level data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 300,
  fig.width = 5, fig.height = 3
)
```

The main examples in _cfr_ use incidence data (i.e. counts of cases and deaths by day). However, in some instances individual-level data will be available, which opens up additional analysis options, but also some potential biases.

::: {.alert .alert-primary}
## Use case {-}

We want to estimate the case fatality risk (CFR) from individual-level linelist data.
:::

::: {.alert .alert-secondary}
### What we have {-}

* A linelist of case onset timings, as well as outcome (i.e. death, recovery) if it has occurred by this point.
* Data on the distribution of delay from onset-to-death, describing the probability an individual will die $t$ days after they were initially infected.
:::

```{r, message = FALSE, warning=FALSE, eval = TRUE}
# load {cfr} and data packages
library(cfr)

# packages to wrangle and plot data
library(dplyr)
library(magrittr)
library(ggplot2)
library(tidyr)
library(incidence2)
```

## Estimation based on cases with known outcomes

If we have individual-level data on case onset timings and outcomes where available, it is common to filter the data to focus on cases only with known outcomes. This means the case fatality risk can then be calculated from this filtered known outcome dataset:
$$
\frac{\text{Total deaths}}{\text{Total deaths}+ \text{Total recoveries}}
$$

One limitation of this method is that it implicity assumes the delay from onset-to-death and onset-to-recovery are the same (see below for the mathematical proof). If these delays are noticeably different, the above calculation will produce a biased estimate of the CFR. For example, if onset-to-recovery is much longer than onset to death, then the denominator above will be an underestimate of the true number of cases that would eventually recover who have similar onset timings to those that would eventually die. Hence this calculation would inflate the CFR estimate. After the mathematical explanation below, the next section will discuss estimation methods based on the expected number of known death outcomes using functionality in _cfr_.

### Mathematical explanation for bias
If we have a linelist with known outcomes and onset-death delay follows the same distribution as onset-recovery, and CFR = $p$, then at the current time,

$$
E(\text{Total deaths to date}) = \sum_{t} p \sum_j f_{j} I_{t-j} 
$$

and

$$
E(\text{Total recoveries to date}) = \sum_{t}(1-p) \sum_j f_{j} I_{t-j} 
$$

where $I_t$ is number of new symptomatic infections on day $t$ and $f_i$ is the probability of outcome $i$ days after symptom onset.

Hence
$$
E(\text{Total deaths})/ [ E(\text{Total deaths})+E(\text{Total recoveries}) ] \\
=[ \sum_{t} p \sum_j f_{j} I_{t-j} ]/[ \sum_{t} p \sum_j f_{j} I_{t-j}  +  \sum_{t}(1-p) \sum_j f_{j} I_{t-j}] \\
=[p \sum_{t}  \sum_j f_{j} I_{t-j} ]/[ \sum_{t}  \sum_j f_{j} I_{t-j} ]   = p
$$

However, if delay to death $f_j^D$ is different to delay to recovery $f_j^R$, we have:

$$
E(\text{Total deaths})/ [ E(\text{Total deaths})+E(\text{Total recoveries}) ]  = \frac{\sum_{t} \sum_j p f_j^D I_{t-j}}{\sum_{t} \sum_j p f_j^D I_{t-j} + \sum_{t} \sum_j (1-p) f_j^R I_{t-j}}
$$
And hence this will not simplify to provide an unbiased estimate of $p$. 

## Estimation based on expected number of known death outcomes

If the delay from onset-to-death and onset-to-recovery are different, one option is to use [survival analysis methods](https://academic.oup.com/aje/article/162/5/479/82647) to estimate relative hazards (i.e. fatality risk) over time.

If we are only interested in an overall estimate of CFR, a simpler alternative is to first calculate the proportion of cases in the linelist that we would expect to have a known outcome by this point if the outcome were fatal. We can then use this estimated proportion with known fatal outcomes to calculate the CFR. This is the calculation performed by `cfr_static()`, and hence this function can give us a better estimate of CFR when delays to death and recovery are not the same.

## Simulated comparison of above methods

To compare these methods, we first simulate some case onset timings and outcomes (i.e. deaths and recoveries)

```{r}
# Simulate data
nn <- 1e3 # Number of cases to simulate
pp <- 0.1 # Assumed CFR
set.seed(10) # Set seed for reproducibility

# Generate random case onset timings in Jan & Feb 2024
case_onsets <- as.Date("2024-01-01") + sample(1:60,nn,replace=T)

# Define current date of data availability (i.e. max follow up)
max_obs <- as.Date("2024-01-20") 

# Sample delays from onset to outcome

# 1. Deaths: assume mean delay = 5 days, sd = 5 days
# log_param <- epiparameter::convert_summary_stats_to_params("lnorm",mean=5,sd=5)
log_param <- c(meanlog=1.262864,sdlog=0.8325546)
delay_death <- function(x){dlnorm(x,meanlog = log_param$meanlog,sdlog = log_param$sdlog)}

outcome_death <- round(rlnorm(round(nn*pp),meanlog = log_param$meanlog,sdlog = log_param$sdlog)) 

# 2. Recoveries: assume mean delay = 15 days, sd = 5 days
# log_param <- epiparameter::convert_summary_stats_to_params("lnorm",mean=15,sd=5)
log_param <- c(meanlog=2.65537,sdlog=0.3245928)
outcome_recovery <- round(rlnorm(round(nn*(1-pp)),meanlog = log_param$meanlog,sdlog = log_param$sdlog)) 

# Create vector of outcome dates
all_outcomes <- case_onsets + c(outcome_death,outcome_recovery)

# Create vector of outcome types
type_outcome <- c(rep("D",length(outcome_death)),rep("R",length(outcome_recovery)))

# Create vector of known outcomes as of max_obs
known_outcomes <- type_outcome; known_outcomes[(all_outcomes>max_obs)] <- ""
```

First, we calculate CFR by filtering to focus only on cases with known outcomes:
```{r}
# Filter on known outcomes
total_deaths <- sum(known_outcomes=="D")
total_outcomes <- (total_deaths+sum(known_outcomes=="R"))

# Calculate CFR with 95% CI
cfr_filter <- binom.test(total_deaths,total_outcomes)
cfr_estimate <- (signif(as.numeric(c(cfr_filter$estimate,cfr_filter$conf.int)),3))
cfr_estimate
```

Next, we calculate CFR based on expected fatal outcomes known to date, using `cfr_static()`
```{r}
# Get times of death for fatal outcomes
death_times <- (case_onsets)[type_outcome=="D"]+outcome_death

# Create a single data.frame with event types
events <- data.frame(
  dates = c(case_onsets, death_times),
  event = c(rep("cases", length(case_onsets)), rep("deaths", length(death_times)))
)

# Use incidence2 to calculate counts of cases and deaths by day
counts <- incidence2::incidence(events, date_index = "dates", groups = "event", complete_dates = TRUE)

# Pivot the incidence object to get a data frame with counts for both cases and deaths
df <- counts %>%
  tidyr::pivot_wider(names_from = event, values_from = count, values_fill = 0) %>%
  dplyr::rename(date=date_index)
  
cfr_static(df,delay_death)
```
