---
title: "Setup and Installation"
output: rmarkdown::html_vignette
bibliography: resources/library.bib
csl: resources/bmj.csl
vignette: >
  %\VignetteIndexEntry{Setup and Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

This short vignette makes sure the key packages are installed properly to use
`datadelay` easily. 

## Essential packages
### Installing `devtools`
We use the `devtools` package to install the required packages: `datadelay`, 
`epiparameter` and (optionally) `covidregionaldata`. To install and load the
`devtools` package, use the following commands:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
install.packages("devtools")
library(devtools)
```


### Installing `datadelay`
`datadelay` can be installed from Github, using the `devtools` package, with the
following command:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
install_github(repo = paste0("epiverse-trace", "/", "datadelay"))
```
### Installing `epiparameter`
The `epiparameter` package is also from the *Epiverse* and has been developed to
work seemlessly with `datadelay` and other *Epiverse* packages. We therefore 
describe it as essential for `datadelay`. However, in the case whereby the user 
supplies the delay distribution, the `epiparameter` package is not needed. We 
install `epiparameter` using the following command:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
install_github(repo = paste0("epiverse-trace", "/", "epiparameter"))
```

## Optional packages — required to run the code in the vignettes
### Installing `covidregionaldata`
The `covidregionaldata` package is the source of the case and death time-series
used in the case studies in the vignettes for the `datadelay` package. 
Therefore, to ensure that the examples in the vignettes run, we install 
`covidregionaldata` using the following command:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
install.packages("covidregionaldata",
                 repos = "https://epiforecasts.r-universe.dev")
```

### Installing `knitr` and `kableExtra`
The `knitr` package builds tables — called `kables`, using the `kable()` 
function — in html or PDF form automatically from data.frames. We use it in the 
vignettes to build the tables. To install it,
run the following command:

```{r, message = FALSE, warning = FALSE, eval = FALSE}
install.packages("covidregionaldata",
                 repos = "https://epiforecasts.r-universe.dev")
```
